---
title: "scCertify: Explainable Confidence Scoring for Single-Cell Annotations"
author: "Jaya Surya Doddetipalli"
date: "`r Sys.Date()`"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{scCertify: Explainable Confidence Scoring for Single-Cell Annotations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

# Introduction

Single-cell RNA sequencing (scRNA-seq) enables the identification of diverse cellular populations from complex biological samples. Numerous computational methods have been developed to assign cell-type labels to individual cells; however, most annotation methods provide little information regarding the reliability of each prediction.

**scCertify** complements existing annotation workflows by estimating confidence for predicted cell identities rather than performing annotation itself. The package combines multiple independent sources of evidence into a calibrated confidence score together with interpretable explanations.

The confidence framework integrates four complementary components:

- Marker gene enrichment
- Neighborhood agreement
- Entropy-based uncertainty
- Doublet information

The resulting confidence scores can be used to identify uncertain annotations, prioritize manual review, improve downstream analyses, and increase the transparency of automated annotation workflows.

# Installation

Install the development version using `BiocManager`.

```{r eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("scCertify")
```

Load the package.

```{r eval=FALSE}
library(scCertify)
```

# Workflow Overview

A typical scCertify workflow consists of:

1. Predict cell identities using an external annotation method.
2. Supply a marker gene database.
3. Compute marker consistency scores.
4. Measure neighborhood agreement.
5. Estimate entropy-based uncertainty.
6. Combine evidence into confidence scores.
7. Interpret confidence using explanation utilities.

The package supports both **Seurat** and **SingleCellExperiment** objects through a common interface.

# Input Requirements

The input object should already contain predicted cell labels.

For Seurat objects:

```{r eval=FALSE}
head(object@meta.data)
```

For SingleCellExperiment objects:

```{r eval=FALSE}
head(SummarizedExperiment::colData(object))
```

The metadata should include a column containing predicted labels (default: `predicted_label`).

# Marker Database

Marker genes are supplied as a named list.

```{r eval=FALSE}
markers <- list(
    "T cell" = c("CD3D","CD3E","TRBC1"),
    "B cell" = c("MS4A1","CD79A","CD79B"),
    "Monocyte" = c("LYZ","S100A8","S100A9")
)
```

# Marker Consistency Score

Marker consistency evaluates whether a cell expresses genes expected for its predicted identity. scCertify uses the UCell algorithm to compute robust rank-based enrichment scores.

```{r eval=FALSE}
marker_scores <- marker_score(
    object,
    markers
)
```

Higher scores indicate stronger molecular support for the assigned annotation.

# Neighborhood Agreement Score

Correctly annotated cells generally occur near cells of the same identity in low-dimensional space.

```{r eval=FALSE}
neighbor_scores <- neighbor_score(
    object
)
```

Neighborhood agreement ranges from 0 to 1, where larger values indicate greater agreement with nearby cells.

# Entropy-Based Uncertainty

Entropy provides a quantitative measure of annotation uncertainty.

```{r eval=FALSE}
object <- entropy_score(
    object
)
```

Low entropy indicates confident predictions, whereas high entropy suggests ambiguous assignments.

# Computing Confidence Scores

The final confidence score integrates all available evidence.

The default weighting scheme is:

| Component | Weight |
|:----------|------:|
| Marker enrichment | 35% |
| Neighborhood agreement | 35% |
| Entropy | 20% |
| Doublet penalty | 10% |

```{r eval=FALSE}
object <- cell_certify(
    object,
    markers
)
```

The resulting metadata include:

- marker_score
- neighbor_score
- confidence_score
- confidence_class

For Seurat objects:

```{r eval=FALSE}
head(
    object@meta.data[
        ,
        c(
            "marker_score",
            "neighbor_score",
            "confidence_score",
            "confidence_class"
        )
    ]
)
```

For SingleCellExperiment objects:

```{r eval=FALSE}
head(
    SummarizedExperiment::colData(object)
)
```

# Custom Confidence Thresholds

Confidence classes can be customized.

```{r eval=FALSE}
object <- cell_certify(
    object,
    markers,
    moderate_threshold = 0.4,
    high_threshold = 0.75
)
```

# Explaining Confidence

Individual cells can be inspected using:

```{r eval=FALSE}
explain_cell(
    object,
    cell_id = "Cell1"
)
```

To understand the major contributors to a confidence score:

```{r eval=FALSE}
explain_confidence(
    object,
    cell_id = "Cell1"
)
```

Possible explanations include:

- Weak marker enrichment
- Neighborhood disagreement
- High entropy
- Possible doublet contamination

# Interpreting Results

High-confidence annotations generally exhibit:

- Strong marker enrichment
- Strong neighborhood agreement
- Low entropy
- Low doublet scores

Low-confidence annotations frequently correspond to:

- Ambiguous transcriptional states
- Rare cell populations
- Doublets
- Poor-quality cells
- Incorrect annotations

These cells may benefit from manual inspection before downstream analyses.

# Compatibility

scCertify supports both:

- Seurat
- SingleCellExperiment

allowing integration into CRAN- and Bioconductor-based workflows.

# Summary

scCertify provides an explainable confidence scoring framework for single-cell annotation. By combining marker enrichment, neighborhood agreement, entropy, and doublet information, it quantifies annotation reliability while providing interpretable explanations for each confidence assignment.

The package is intended to complement existing annotation pipelines and assist with quality control, annotation refinement, downstream filtering, and biological interpretation.

# References

- Andreatta M, Carmona SJ (2021). *UCell: Robust and scalable single-cell gene signature scoring*. Computational and Structural Biotechnology Journal.

- Aran D, Looney AP, Liu L, et al. (2019). *Reference-based analysis of lung single-cell sequencing reveals a transitional profibrotic macrophage*. Nature Immunology.

# Session Information

```{r session-info}
sessionInfo()
```
